Local Type Inference for Polarised System F with Existentials

### 2024-01-23 oopsla24/paper.tex
WIP: Declarative
Restructure proofs
wip: extensions
wip

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  {\color{red}CHANGE!!}
  This paper addresses the challenging problem of type inference for
  Impredicative System F with existential types, a critical aspect of many
  programming languages. While System F serves as the basis for type systems in
  numerous languages, existing type inference techniques for Impredicative
  System F are undecidable due to the presence of existential ($[[∃]]$) 
  and polymorphic ($[[∀]]$) types. Consequently, current algorithms are often ad-hoc and
  sub-optimal. This paper presents novel contributions in the form of a local
  type inference algorithm for Impredicative System F with existential types.
  The algorithm introduces innovative techniques, such as a unique combination
  of unification and anti-unification, a full correctness proof, and the use of
  control structures inspired by \CBPV. Additionally, the paper discusses a type
  inference framework that allows the algorithm to be applied to different type
  systems, offering insights into the under-researched area of impredicative
  existential type inference. 
\end{abstract}

@@

\section{Algorithm}
\input{\genDir/tex/04-algorithm.tex}

\section{Proof}
\input{\genDir/tex/05-proof.tex}

\section{Extensions}
\input{\genDir/tex/06-extensions.tex}

@@

\cite{botlanMLFRaisingML2003}
\cite{dunfieldBidirectionalTyping2020}

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  % {\color{red}CHANGE!!}
  % This paper addresses the challenging problem of type inference for
  % Impredicative System F with existential types, a critical aspect of many
  % programming languages. While System F serves as the basis for type systems in
  % numerous languages, existing type inference techniques for Impredicative
  % System F are undecidable due to the presence of existential ($[[∃]]$) 
  % and polymorphic ($[[∀]]$) types. Consequently, current algorithms are often ad-hoc and
  % sub-optimal. This paper presents novel contributions in the form of a local
  % type inference algorithm for Impredicative System F with existential types.
  % The algorithm introduces innovative techniques, such as a unique combination
  % of unification and anti-unification, a full correctness proof, and the use of
  % control structures inspired by \CBPV. Additionally, the paper discusses a type
  % inference framework that allows the algorithm to be applied to different type
  % systems, offering insights into the under-researched area of impredicative
  % existential type inference. 
\end{abstract}

@@

\section{The Algorithm}
\input{\genDir/tex/04-algorithm.tex}

\section{Algorithm Correctness}
\input{\genDir/tex/05-proof.tex}

\section{Extensions and Future Work}
\input{\genDir/tex/06-extensions.tex}

@@

### 2024-01-23 oopsla24/tex/02-overview.tex
wip: correctness
wip

@@

@@

@@

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a (positive) term variable; 
        \ilyam{why no negatives? Following CBPV}
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ operate similarly to 
        the bind of a monad: they take a suspended computation $[[v]]$,
        apply it to a list of arguments, bind the result 
        (which is expected to be pure) to a variable $[[x]]$,
        and continue with a computation $[[c]]$.
        If the resulting type of the application is unique, 
        one can omit the type annotation, as in the second form:
        it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

\paragraph{Missing constructors}
Notice that the language does not have first-class applications: 
their role is played by the applicative let forms, binding 
the result of a \emph{fully applied} function to a variable.
Also notice that the language does not have a type application 
(i.e. the eliminator of $[[∀]]$) and dually, 
it does not have \pack (i.e. the constructor of $[[∃]]$).
This is because the instantiation of polymorphic and existential 
types is inferred by the algorithm.  \ilyam{refer to the extension chapter}

>>>>

\subsection{Translation to \fexists}

The language of \fexists is different from standard \systemf:
every term and type has either positive or negative polarity.
The type variables are annotated with their polarities
(e.g. $[[α⁺]]$ or $[[β⁻]]$), and the types can change polarity
via the shift operators $[[↑]]$ (positive to negative) and
$[[↓]]$ (negative to positive).

To establish a correspondence between \fexists and standard unpolarized systems,
we present a translation between \fexists and \systemf. The translation of
\systemf to \fexists is given in \cref{fig:translation}.

\begin{figure}
  \begin{equation*}
  \begin{aligned}
    [[|↓iN|]] &= [[|iN|]] \\
    [[|↑iP|]] &= [[|iP|]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[|α±|]] &= [[a]] \\
    [[|iP → iN|]] &= [[|iP| → |iN|]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
    [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
  \end{aligned}
  \end{equation*}
  \caption{Type depolarization}
  \label{fig:depolarization}
\end{figure}

The translation from \systemf to \fexists can be done in two canonical ways
depending on the desired polarity. The polarity will also determine the 
execution strategy \ilyam{how?}.
Moreover, the translation procedure depends on the \emph{polarity} of the
free type variables. Thus, the translation functions 
$[[-| fT |-Γ]]$ and $[[+| fT |+Γ]]$ are indexed by a context $[[Γ]]$, 
containing either $[[α⁺]]$ or $[[α⁻]]$ for each variable $[[α]]$ free in $[[fT]]$.

\begin{figure}
  \begin{equation*}
  \begin{aligned}
    [[-| a |-Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
    [[-| a |-Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
    [[-| fA → fB |-Γ]] &= [[ +|fA|+Γ → -|fB|-Γ  ]] \\
    [[-| ∀α.fT |-Γ]] &= [[ ∀α⁺.-|fT|-Γ,α⁺ ]] \\
    [[-| ∃α.fT |-Γ]] &= [[ ↑ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[+| a |+Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
    [[+| a |+Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
    [[+| fA → fB |+Γ]] &= [[ ↓ +|fA|+Γ → -|fB|-Γ  ]] \\
    [[+| ∀α.fT |+Γ]] &= [[ ↓∀α⁺.-|fT|-Γ,α⁺ ]] \\
    [[+| ∃α.fT |+Γ]] &= [[ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
  \end{aligned}
  \end{equation*}
  \caption{Type polarization}
  \label{fig:polarization}
\end{figure}

@@

Suppose that
we have the following variables in context $[[Φ]]$:

\begin{align*}
  [[map]] &: [[↓∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺]] \\
  [[len]] &: [[↓∀α⁺.List α⁺ → ↑Int]] \\
  [[choose]] &: [[↓∀α⁺.α⁺ → α⁺ → ↑α⁺]] \\
  [[id]] &: [[↓∀α⁺.α⁺ → ↑α⁺]] \\
  [[auto]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
\end{align*}

Then

\begin{align*}
  &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
  &[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]] \ilyam{(no)}\\
\end{align*}

@@

\paragraph{Type Context and Type Well-formedness}

In the construction of \fexists, a type context (denoted as $[[Γ]]$) is
represented as a \emph{set} of positive and negative type variables and it is used to
assert the well-formedness of types. The well-formedness of a type is denoted as
$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ and it asserts that all type variables are either
bound by a quantifier ($[[∀]]$ and $[[∃]]$) or declared in the context $[[Γ]]$.
The well-formedness checking is an \emph{algorithmic} procedure. As commonly
done, we represent it as a system of inference rules, that correspond to a
recursive algorithm taking the context and the type as input.

@@

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a term variable.
      Following the \CBPV stratification, we only have \emph{positive} (value)
      term variables;
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ operate similarly to 
        the bind of a monad: they take a suspended computation $[[v]]$,
        apply it to a list of arguments, bind the result 
        (which is expected to be pure) to a variable $[[x]]$,
        and continue with a computation $[[c]]$.
        If the resulting type of the application is unique, 
        one can omit the type annotation, as in the second form:
        it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

\paragraph{Missing constructors}
Notice that the language does not have first-class applications: 
their role is played by the applicative let forms, binding 
the result of a \emph{fully applied} function to a variable.
Also notice that the language does not have a type application (i.e. the eliminator of $[[∀]]$) and dually, it does not have \pack (i.e. the constructor of $[[∃]]$).
This is because the instantiation of polymorphic and existential types is inferred by the algorithm. 
In \cref{sec:extensions}, we discuss the way to modify the system to introduce \emph{explicit} type applications.

### 2024-01-23 oopsla24/tex/03-declarative-system.tex
wip

\paragraph{Invariant Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, after certain changes 
\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form,
 as we will discuss in \cref{todo}.

@@

A property that is important for the subtyping algorithm, 
in particular for the type \emph{upgrade} procedure (\cref{todo}),
is the preservation of free variables by subtyping.
Informally, it says that the free variables
of a positive type cannot disappear in its subtypes,
and the free variables of a negative type
cannot disappear in its supertypes.

@@

The normalization algorithm is shown in \cref{fig:type-nf}.
The steps (i-ii) are implemented by the ordering function, which takes a set of variables $[[varset]]$
and a type and returns a list of variables from $[[varset]]$ that occur in the type in the order of their
first occurrence. Its formal definition can be found in \cref{todo}.

@@

\paragraph{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

\paragraph{Applicative Let Binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.
  The difference is that the former rule is for the 
  \emph{unannotated} let binder, and thus
  it requires the resulting type of application to be unique
  (up to equivalence), so that the type of the bound variable $[[x]]$
  is known before it is put into the context.
  The latter rule is for the \emph{annotated} binder,
  and thus, the type of the bound $[[x]]$ is given, 
  however, the rule must check that this type is a
  a supertype of the inferred type of the application. 
  This check is done by invoking the subtyping judgment
  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
  This judgment is more restrictive than checking bare 
  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
  to make the algorithm complete as it allows us to preserve
  certain invariants (see \cref{todo}).
  In \cref{todo} we discuss how this restriction can be relaxed
  together with invariant shift subtyping.

@@

\paragraph{Application of a Polymorphic Type $[[∀]]$}
  The complexity of the system is hidden in the rules, 
  whose output type is not immediately defined by their input
  and the output of their premises 
  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
  the instantiation of the quantifying variables is
  not known a priori. The algorithm we present in \cref{todo} 
  delays this instantiation until more information about it 
  (in particular, typing constraints) is collected.

@@

\paragraph{Application of an Arrow Type}
  Another important application rule is \ruleref{\ottdruleDTArrowAppLabel}.
  This is where the subtyping is used to check that the type of the argument
  is convertible to (a subtype of) the type of the function parameter.
  In the algorithm (\cref{todo}), this subtyping check will provide the constraints
  we need to resolve the delayed instantiations of the quantifying variables.

>>>>

\paragraph{Invariant Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, after certain changes 
\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form,
thereby increasing the expressiveness of the system. These
changes are discussed in \cref{sec:weakening-invariant}.

@@

A property that is important for the subtyping algorithm, 
in particular for the type \emph{upgrade} procedure (\cref{sec:lub}),
is the preservation of free variables by subtyping.
Informally, it says that the free variables
of a positive type cannot disappear in its subtypes,
and the free variables of a negative type
cannot disappear in its supertypes.

@@

The normalization algorithm is shown in \cref{fig:type-nf}.
The steps (i-ii) are implemented by the \emph{ordering} function
$[[ord varset in iN]]$ and $[[ord varset in iP]]$.
For a set of variables $[[varset]]$, and a type, it 
returns a list of variables from $[[varset]]$ that occur in the type in 
the order of their first occurrence.
For brevity, we omit the formal definition of ordering,
referring the reader to the appendix.

@@

\paragraph{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

\paragraph{Applicative Let Binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.
  The difference is that the former rule is for the 
  \emph{unannotated} let binder, and thus
  it requires the resulting type of application to be unique
  (up to equivalence), so that the type of the bound variable $[[x]]$
  is known before it is put into the context.
  The latter rule is for the \emph{annotated} binder,
  and thus, the type of the bound $[[x]]$ is given, 
  however, the rule must check that this type is a
  a supertype of the inferred type of the application. 
  This check is done by invoking the subtyping judgment
  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
  This judgment is more restrictive than checking bare 
  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
  to make the algorithm complete as it allows us to preserve
  certain invariants throughout the algorithm to ensure
  the decidability of the produced unification task.
  In \cref{sec:proof-subtyping}, we discuss these invariants in
  more detail, and in \cref{sec:weakening-invariant} we suggest 
  a way to relax them.

@@

\paragraph{Application of a Polymorphic Type $[[∀]]$}
  The complexity of the system is hidden in the rules, 
  whose output type is not immediately defined by their input
  and the output of their premises 
  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
  the instantiation of the quantifying variables is
  not known a priori. The algorithm we present in \cref{sec:algorithm} 
  delays this instantiation until more information about it 
  (in particular, typing constraints) is collected.

@@

\paragraph{Application of an Arrow Type}
  Another important application rule is \ruleref{\ottdruleDTArrowAppLabel}.
  This is where the subtyping is used to check that the type of the argument
  is convertible to (a subtype of) the type of the function parameter.
  In the algorithm (\cref{sec:algorithm}), this subtyping check will provide the constraints
  we need to resolve the delayed instantiations of the quantifying variables.
